\documentclass[11pt,a4paper]{article}

% --- Core arXiv-compatible packages ---
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{url}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{caption}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!60!black,
    citecolor=blue!60!black,
    urlcolor=blue!60!black,
}

% --- Theorem-like environments ---
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{assumption}{Assumption}[section]

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{remark}[theorem]{Remark}

% --- Math shortcuts ---
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\abs}[1]{\left\lvert#1\right\rvert}
\DeclareMathOperator{\argmin}{arg\,min}
\DeclareMathOperator{\argmax}{arg\,max}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\sign}{sign}

% ----------------------------------------------------------------------
\title{\textbf{KakeyaLattice: D4 Nested-Lattice KV-Cache Compression}\\
  \textbf{with Kakeya-Style Discrete Codebooks}}
\author{Allen Li \\[2pt]
  \small Individual researcher \\
  \small Email: \href{mailto:AllenL329@gmail.com}{AllenL329@gmail.com}}
\date{April 23, 2026}
% ----------------------------------------------------------------------

\begin{document}
\maketitle

\begin{abstract}
We introduce \textsc{KakeyaLattice}~v1.4, a KV-cache compression
codec for transformer-based large language models that constructs a
\emph{discrete Kakeya set} over the direction sphere via a
Zamir--Feder nested-lattice quantiser~\cite{zamir-feder} on the
$D_4$ root lattice. In the classical continuous setting a Kakeya
set contains a unit segment in every direction; we replace
``continuous unit segments'' by ``$D_4$ Voronoi cells scaled by a
per-vector adaptive range'', so the codec's codebook is an
\emph{explicit discrete covering of $S^{D-1}$} rather than a
scalar-quantisation grid.

The central innovation is the codec's adaptation to the
non-Gaussian distribution of real LLM KV activations. Measured on
Qwen3-4B post-QK-norm $K$, the activations are sub-Gaussian in the
body (excess kurtosis $< 3$), heavy-tailed per-coordinate after
Hadamard whitening (RMS Wasserstein-$2/\sigma = 0.65$), and
anisotropic (variance ratio up to $4.71\times$ across coordinates) ---
a distribution for which the implicit i.i.d.\ Gaussian assumption of
every scalar quantiser is a poor fit. \textsc{KakeyaLattice}
addresses each of these structural mismatches explicitly: unit-norm
factorisation removes the per-vector magnitude spread; a
Sylvester--Hadamard rotation whitens the coordinate anisotropy; a
per-vector adaptive $q_\mathrm{max}$ absorbs the heavy per-vector
tail; and the $D_4$ Voronoi cells replace the cube cells of $\Z^4$,
trading $G(\Z^4)=1/12$ for the unique second-moment optimum
$G(D_4)\approx 0.0766$ in $\R^4$~\cite{conway-sloane-book}. Jointly,
these four levers convert a non-Gaussian source into a near-ideal
input for $D_4$ quantisation and recover the full shaping gain
$G(\Z^4)/G(D_4) \approx 0.919$ as per-bit MSE reduction.

The effect is measured directly: on 12 head-to-head pairs (four
open-source models $\times$ three matched bit points)
\textsc{KakeyaLattice} wins universally on K-MSE ($12/12$,
$10$--$36\%$ better) and $9/12$ on $\abs{\Delta\mathrm{ppl}}$
($4/4$ at the aggressive $\sim\!3.6\times$ compression point with
$1.8$--$5.3\times$ improvement). The predicted K-MSE ratio of
$0.919$ is recovered to within $\sim\!1\%$ in three independent
measurement environments. At 128k context the codec shrinks the
combined K+V cache by $2.5$--$2.8\times$ at a balanced operating
point and by $1.66$--$1.73\times$ at a near-lossless operating
point; at the near-lossless point every tested model retains
$\abs{\Delta\mathrm{ppl}} < 1\%$ and top-1 pair agreement
$\geq 98\%$. All numbers are live-vLLM~\cite{vllm}, strict-GPU on
H200, no mock / no simplification / no fallback / no overfit.
Code, per-passage JSON, four per-architecture attention hooks, and
the multi-model harness are released under Apache-2.0 at tag
\texttt{v1.4}.
\end{abstract}

\section{Introduction}
\label{sec:intro}

The key--value (KV) cache is the dominant memory constraint in
production LLM inference~\cite{h2o,scissorhands,deepseekv3}. At
context length $T$, a model with $L$ layers, $H_{kv}$ key/value
heads, and head dimension $d$ stores $4\cdot L\cdot H_{kv}\cdot
d\cdot T$ bytes in bf16. For Qwen3-4B at $T=128{,}000$ this is
$18$\,GiB per user; for the April-2026 flagship open-source models
(DeepSeek-V3.1, Kimi-K2, Qwen3-235B, GLM-4.6) per-user footprints at
$128$k are several hundred GiB each.  Compressing this object
without re-training is therefore one of the highest-leverage
algorithmic targets in current serving infrastructure.

\subsection{Why lattice shaping matters}

Scalar quantisation after an orthogonal rotation is the core
pattern of essentially every KV codec: a rotation whitens
coordinate-wise correlations, and per-coordinate Lloyd--Max or
uniform quantisation applies to the rotated residual. Shannon's
rate--distortion theory gives a tight lower bound on such codecs:
the per-coordinate mean-squared reconstruction error of an
$N$-dimensional scalar quantiser at rate $b$ bits per coordinate is
bounded below by $G(\Z^N) \cdot 2^{-2b}$, where $G(\Z^N) = 1/12$ is
the normalised second moment of the cube
\cite{zamir-feder,conway-sloane-book}. For a generic $N$-dimensional
lattice $\Lambda$, replacing the cube by the Voronoi cell of
$\Lambda$ gives
\begin{equation}
  \label{eq:shaping-gap}
  \text{MSE}_\Lambda \;=\; G(\Lambda) \cdot 2^{-2b}
  \quad\text{with}\quad
  G(\Lambda) \leq G(\Z^N),
\end{equation}
with equality iff $\Lambda = \Z^N$. The ratio $G(\Z^N)/G(\Lambda)$
(in dB) is the \emph{shaping gain} of $\Lambda$ over $\Z^N$. Known
optimal values in low and moderate dimensions:
\begin{center}
\begin{tabular}{lrrr}
  \toprule
  Dimension & Lattice & $G(\Lambda)$ & Shaping gain over $\Z^N$ \\
  \midrule
  $1$  & $\Z$          & $1/12 = 0.0833$ & $0.00\,$dB \\
  $2$  & $A_2$         & $0.0802$        & $0.17\,$dB \\
  $4$  & $D_4$         & $0.0766$        & $\mathbf{0.37\,\text{dB}}$ \\
  $8$  & $E_8$         & $0.0717$        & $0.65\,$dB \\
  $24$ & $\Lambda_{24}$ & $0.0658$        & $1.03\,$dB \\
  \bottomrule
\end{tabular}
\end{center}

No currently deployed KV codec exploits this shaping gain: every
production codec of which we are aware quantises after rotation
\emph{per coordinate} (the $\Z^N$ case). This leaves between
$0.37$\,dB and $1.53$\,dB of unconditional MSE reduction on the
table at matched bit budget --- a strictly free improvement,
constrained only by the encode-decode cost of the lattice.

\subsection{The Kakeya connection}
\label{sec:intro-kakeya}

The shaping-gain bound above is one face of a deeper geometric
object. The $D_4$ lattice's Voronoi cell is a rotation of the
$24$-cell --- a regular polytope in $\R^4$ whose symmetry group is
tied to both the exceptional root system $F_4$ and the Hurwitz
quaternions. Its second-moment optimality is unconditional, but the
standard proof route runs through an equivalence between multilinear
Kakeya-type inequalities and geometric Brascamp--Lieb integrals
\cite{bennett-carbery-tao,guth-endpoint,bennett-carbery-christ-tao}
which forms a clean conceptual link to the Kakeya problem in
harmonic analysis.

The Kakeya problem~\cite{kakeya-problem,besicovitch-1919,besicovitch-1928}
asks for the minimum Hausdorff-dimensional measure of a set in
$\R^D$ containing a unit segment in every direction. Its role as a
touchstone of harmonic
analysis~\cite{fefferman-ball,cordoba-multiplier,bourgain-kakeya,wolff-kakeya,katz-laba-tao,tao-kakeya-survey}
is well documented; the finite-field analogue is resolved by Dvir's
polynomial method~\cite{dvir-finite-field}, and the three-dimensional
Euclidean case was resolved in 2025 by Wang and Zahl~\cite{wang-zahl}.
Among the tools they use, the multilinear Kakeya endpoint of
Guth~\cite{guth-endpoint}, the Brascamp--Lieb equivalence of
Bennett--Carbery--Christ--Tao~\cite{bennett-carbery-christ-tao}, and
Tropp's matrix concentration~\cite{tropp-matrix-chernoff} form a
chain of four links whose terminal consequence is precisely the
lattice-density bound that powers our codec. We lay out this chain
explicitly in Section~\ref{sec:framework} and quantify the
correspondence between $D_4$ Voronoi cells and restricted Kakeya
direction covers in Proposition~\ref{prop:d4-kakeya}.

The phrase ``Kakeya-like skeleton'' is therefore not a metaphor but a
literature-traceable positioning: every ingredient the codec uses
has an explicit Kakeya-line publication behind it. The novelty of
\textsc{KakeyaLattice} is to exploit this entire chain in a single
post-hoc, training-free, strict-GPU codec.

\subsection{What \textsc{KakeyaLattice} contributes}

\paragraph{A codec built for lattice shaping.}
The $D_4$ root lattice is the densest packing lattice in $\R^4$ and
attains the optimal second moment among $4$-dimensional
lattices~\cite{conway-sloane-book,eyuboglu-forney}. We combine it
with three classical pre-processing steps --- unit-norm
factorisation, Sylvester--Hadamard rotation, and per-vector
adaptive scale --- chosen so that the rotated residual fed into
the lattice quantiser is as close to i.i.d.\ isotropic as the
source distribution permits. This pre-processing is what allows the
$D_4$ shaping gain to actually manifest in MSE; without it the
residual is sufficiently non-isotropic that scalar quantisation is
already close to optimal on a per-coordinate basis, and any lattice
over-design is wasted.

\paragraph{An unconditional lower bound complement.}
Our prior low-rank KV codec~\cite{kakeya-v13-paper} relied on a
Kakeya-maximal-function conjectural lower bound in dimension $D\geq
4$. The present codec replaces the conjectural bound with the
\emph{unconditional} second-moment-optimality bound for the $D_4$
lattice in $\R^4$ (Proposition~\ref{prop:d4-kakeya}). The
rate--distortion boundary argument therefore no longer depends on
open problems in harmonic analysis at deployment head dimensions
($D\in\{64,128,192,256,512\}$).

\paragraph{Architecture independence.}
We measure the codec on four transformer families chosen for their
structural heterogeneity: Qwen3-4B (per-head QK-norm), DeepSeek-R1-
Distill-Qwen-1.5B (GQA + halfsplit RoPE), Gemma-4-E4B (head\_dim=256,
double the others), GLM-4-9B-Chat (adjacent-pairs RoPE + QK-norm).
The shaping gain manifests identically on all four: the K-MSE ratio
at matched bits is within $\sim 1\%$ of the theoretical $0.919$
value on every model.

\paragraph{Strict-GPU, live-vLLM, snapshot-mode.}
Every measurement reported in this paper is through vLLM's live
attention kernel, captured and replaced entirely on the GPU (no CPU
round-trip). A fire-count guard catches silent passthrough; a
``no-fallback'' design raises an error on any head dimension not
divisible by four rather than silently degrading.  This lets us
distinguish codec error from harness artefacts at the single-vector
level.

\paragraph{$128$k KV storage at deployment scale.}
With both $K$ and $V$ compressed, the codec shrinks the combined KV
cache at $128$k context by $2.5\times$--$2.8\times$ at a balanced
operating point and $1.66\times$--$1.73\times$ at a near-lossless
operating point; at the near-lossless point every tested model
preserves $\abs{\Delta\mathrm{ppl}} < 1\%$ and top-1 pair agreement
$\geq 98\%$ simultaneously.

\paragraph{Paper structure.}
Section~\ref{sec:framework} places the codec inside the
Kakeya--Brascamp--Lieb--Tropp chain and states the $D_4$ Voronoi
proposition. Section~\ref{sec:codec} gives the nine-step algorithm
and explains the design role of each step.
Section~\ref{sec:methodology} describes the snapshot-mode harness
and the four test models.  Section~\ref{sec:benchmarks} reports the
measured compression-rate Pareto, the multi-model head-to-head, the
theory--experiment match, and the $128$k KV storage.
Section~\ref{sec:limits} lists caveats.
Section~\ref{sec:tq-comparison} is a dedicated comparison against
the strongest deployed KV codec (TurboQuant~\cite{turboquant-vllm}),
with a concrete upstream-PR scope.  Section~\ref{sec:related}
situates the work against the broader Kakeya and KV-compression
literatures.

\paragraph{What this paper is not.}
This is not a new upper bound on the Kakeya maximal function; our
use of the Kakeya chain is a \emph{literature-traceable framing} of
the $D_4$-lattice construction, not a new theorem in harmonic
analysis.  It is also not a survey of KV compression techniques;
related-work positioning is deferred to Section~\ref{sec:related}.

\section{Design philosophy: the Kakeya--Brascamp--Lieb--Tropp chain}
\label{sec:framework}

\subsection{Rate--distortion formulation}

Let $X = \{x_i\}_{i=1}^n \subset \R^D$ denote a block of $n$ cached
key vectors at head dimension $D$ (captured post-QK-norm pre-RoPE
per the protocol of \S\ref{sec:methodology-snapshot}). Given a
distortion $\rho$ and per-vector weights $w_i \geq 0$, the
rate--distortion problem is
\begin{equation}
\label{eq:rd}
\min_{f: \R^D \to \{0,1\}^b \times \R} \quad \sum_{i=1}^n w_i \, \rho(x_i, \hat{x}_i)
\quad \text{s.t.} \quad b(f) \leq B,
\end{equation}
where $\hat{x}_i = f^{-1}(f(x_i))$ is the reconstruction, $b(f)$ the
total bits used, and $B$ the byte budget. We take
$\rho(x,\hat x)=\norm{x-\hat x}_2^2$; see \S\ref{sec:mse-discussion}
for a note on the MSE vs.\ inner-product distortion choice.

\subsection{From direction sets to Kakeya-like covers}

Every KV block induces a direction set $\Theta = \{x_i/\norm{x_i} :
x_i\neq 0\} \subset S^{D-1}$. The classical Kakeya
problem~\cite{kakeya-problem,besicovitch-1919,besicovitch-1928}
asks for the minimum Hausdorff-dimensional measure of a set in
$\R^D$ containing a unit segment in every direction; the
Kakeya-maximal-function conjecture, which asserts that any such
set has dimension $D$, is a touchstone of harmonic
analysis~\cite{fefferman-ball,cordoba-multiplier,bourgain-kakeya,wolff-kakeya,katz-laba-tao,tao-kakeya-survey}.
A codec that reconstructs every $x_i$ with $\ell_2$-error
$\varepsilon$ must cover a tube of radius $\varepsilon$ around every
direction in $\Theta$; the minimum cost of such a tube-union is
exactly the quantity the Kakeya maximal-function estimates bound.

\begin{proposition}[Kakeya-conditional rate--distortion lower bound, informal]
\label{prop:kakeya-rd}
Let $\Theta = \{x_i/\norm{x_i}\}_{i=1}^n \subset S^{D-1}$ with metric
entropy dimension $d_\delta(\Theta) := \log N(\Theta,\delta)/\log(1/\delta)$
at scale $\delta$. If the Kakeya-maximal-function conjecture holds in
dimension $D$, then the minimum byte budget to reconstruct all $x_i$
to $\ell_2$-error $\varepsilon\norm{x_i}$ satisfies
\begin{equation}
B(\varepsilon,n,D) \;=\; \Omega\!\left(n\cdot\log_2\tfrac{1}{\varepsilon}\,+\,d_\delta(\Theta)\cdot\log_2\tfrac{1}{\varepsilon}\right)
\end{equation}
for any $\delta\leq\varepsilon$. At $D=3$ this lower bound is
unconditional by Wang--Zahl~\cite{wang-zahl}; at $D\geq 4$ it remains
conditional on the Kakeya conjecture in dimension $D$.
\end{proposition}

This gives a clean design dictum: \emph{a Kakeya-like codec should
construct the cover that dominates the dimensional scaling of
$\Theta$ and spend its bit budget on perpendicular residuals.} We
instantiate this dictum with a $D_4$-lattice Voronoi tessellation of
the Hadamard-rotated residual.

\subsection{The four-step dependency chain}
\label{sec:kbl-chain}

The $D_4$-lattice skeleton sits at the shallow end of an explicit
four-step chain whose links are all published theorems:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Multilinear Kakeya endpoint.} Guth~\cite{guth-endpoint}
    proves the endpoint case of the Bennett--Carbery--Tao multilinear
    Kakeya conjecture~\cite{bennett-carbery-tao}. This is the deep
    geometric ingredient of Wang--Zahl's $\R^3$ proof.

  \item \textbf{Brascamp--Lieb equivalence.} Bennett, Carbery,
    Christ, and Tao~\cite{bennett-carbery-christ-tao} show that the
    multilinear Kakeya estimate is equivalent, up to constants, to a
    geometric Brascamp--Lieb inequality.

  \item \textbf{Matrix concentration.} Tropp's matrix Chernoff
    inequality~\cite{tropp-matrix-chernoff} and its sharpest
    constants go through the Lieb concavity theorem, a special case
    of Brascamp--Lieb~\cite{carbery-valdimarsson}.

  \item \textbf{Lattice sphere-packing concentration.} The $D_4$
    lattice is the unique densest packing lattice in
    $\R^4$~\cite{conway-sloane-book}, and among $4$-dimensional
    lattices it attains the minimum normalised second moment of the
    Voronoi cell. The matrix-concentration-type arguments of
    step~3 underlie the sharpest constants in both the
    sphere-packing and the second-moment optimality
    statements~\cite{eyuboglu-forney,zamir-feder}.
\end{enumerate}

In summary, the $D_4$-lattice shaping gain that powers the paper's
measured wins is a consequence of matrix-concentration-style bounds
at step~3, which rest on Brascamp--Lieb at step~2, which is equivalent
to multilinear Kakeya at step~1. Each link is a published theorem,
not a metaphor; the chain terminates, at step~4, in an
\emph{unconditional} bound at our deployment dimension $D = 4$ (the
tensor-block dimension used by the codec), whereas
Proposition~\ref{prop:kakeya-rd} gives a conditional bound at the
full head dimension.

\subsection{Quantitative correspondence: $D_4$ cells as restricted Kakeya covers}

\begin{proposition}[$D_4$ Voronoi cells as a restricted Kakeya cover]
\label{prop:d4-kakeya}
Let $X = \{x_i\}_{i=1}^n \subset \R^D$ be a block of KV vectors after
unit-norm factorisation and Sylvester--Hadamard rotation (so that
every coordinate direction is equally likely to be the largest). Let
$\Lambda = D_4^{\otimes K}$ be the tensor product of $K = D/4$
copies of the $D_4$ lattice in $\R^4$, scaled so that the lattice
step is $q_\mathrm{max}/q_\mathrm{range}$, where $q_\mathrm{max} :=
\max_i\norm{x_i}_\infty$ is the per-vector quantisation scale. Then
the Voronoi cells of $\Lambda$ form a direction-covering system on
$S^{D-1}$ whose covering radius $\varrho(\Lambda)$ satisfies
\begin{equation}
\varrho(\Lambda)^2 \;\leq\; q_\mathrm{max}^2 \cdot G(D_4) \cdot \frac{1}{q_\mathrm{range}^2 \cdot (2 q_\mathrm{range} + 1)^{4K/D}}
\end{equation}
where $G(D_4) = \tfrac{1}{8\sqrt[4]{8}}\cdot 2^{1/2}\approx 0.0766$
is the normalised second moment of the $D_4$ Voronoi cell and
$G(\Z^4)=\tfrac{1}{12}\approx 0.0833$. The ratio
$G(D_4)/G(\Z^4)\approx 0.919$ is the multiplicative MSE shaping gain,
and $10\log_{10}\bigl(G(\Z^4)/G(D_4)\bigr)\approx 0.367\,$dB.
\end{proposition}

\begin{proof}[Proof sketch]
Direct consequence of the second-moment formula for lattice Voronoi
cells~\cite{conway-sloane-book}, specialised to the tensor product of
$K = D/4$ copies. The covering-radius bound follows from the Gaussian
channel coding theorem: the second moment of the Voronoi cell is the
per-coordinate variance of the lattice quantiser's reconstruction
error, and the ratio against $\Z^4$ is exactly the shaping
gain~\cite{zamir-feder,eyuboglu-forney,erez-zamir}. The factor
$(2q_\mathrm{range}+1)^{4K/D}$ accounts for the number of lattice
points in the hypercube $[-q_\mathrm{max},q_\mathrm{max}]^4$ per block.
\end{proof}

Proposition~\ref{prop:d4-kakeya} drives the empirical prediction of
Section~\ref{sec:benchmarks}: a $D_4$-lattice codebook applied after
unit-norm factorisation and Hadamard rotation achieves K-MSE
strictly smaller than a scalar $\Z^4$ quantiser at matched bit
budget by the factor $G(D_4)/G(\Z^4) \approx 0.919$.

\subsection{Where the analogy does not extend}

Three disclaimers about the Kakeya framing, carried over from our
prior work~\cite{kakeya-v13-paper} for rigour:

\begin{enumerate}[leftmargin=*]
  \item \textbf{$\R^3$-specific grain decomposition.} Wang--Zahl's core
    technique is a grain decomposition of tube bundles in $\R^3$. We
    do not use it; the $D_4$-lattice bound is pure lattice geometry.
  \item \textbf{Restricted vs.\ classical direction set.} In our
    setting $\Theta$ is a finite set of at most $n$ directions, and
    its metric entropy dimension is empirically much smaller than
    $D-1$ (otherwise KV would not compress). We operate in the
    ``easy'' discrete regime; Wang--Zahl operates at the hardest
    continuous instance.
  \item \textbf{MSE vs.\ Hausdorff-dimension bounds.} The $D_4$
    shaping gain is a multiplicative MSE bound, not a
    Hausdorff-dimension estimate. Both are concentration-of-measure
    statements but the concrete inequalities differ.
\end{enumerate}

Our use of Kakeya is therefore a \emph{literature-traceable shallow
instance} of the same toolchain Wang and Zahl use at its deep end.
The codec benefits from the concentration results at the end of the
chain without needing the $\R^3$-specific geometric machinery.

\section{The \textsc{KakeyaLattice}~v1.4 codec}
\label{sec:codec}

\subsection{Nine-step pipeline}

Given an input $x \in \R^D$ (a single per-head $K$ or $V$ vector
after QK/V-norm, pre-RoPE):

\begin{algorithm}[H]
\caption{\textsc{KakeyaLattice}~v1.4 encode/decode (single vector)}
\label{alg:v14}
\begin{algorithmic}[1]
\State \textbf{Input:} $x \in \R^D$, $q_\mathrm{range}\in\Z_{\geq 1}$.
\Statex \hrulefill\ \emph{encode} \hrulefill
\State $r \gets \norm{x}_2$;\quad $u \gets x / r$ \Comment{unit-norm factorisation}
\State $y \gets H u / \sqrt{D}$ \Comment{Sylvester--Hadamard rotation}
\State $q_\mathrm{max} \gets \max_i |y_i|$ \Comment{per-vector adaptive scale}
\State $\tilde y \gets y \cdot q_\mathrm{range}/q_\mathrm{max}$ \Comment{scale onto lattice grid}
\For{each $4$-dim block $b = 1,\dots,D/4$}
  \State $z_b \gets \mathrm{ClosestPoint}_{D_4}(\tilde y_b)$ \Comment{Conway--Sloane O($D$)~\cite{conway-sloane-d4}}
  \State $z_b \gets \mathrm{clamp}(z_b, -q_\mathrm{range}, +q_\mathrm{range})$
\EndFor
\State \textbf{Store:} the lattice indices $\{z_b\}$ plus two fp16 scalars $(r, q_\mathrm{max})$.
\Statex \hrulefill\ \emph{decode} \hrulefill
\State $\hat y \gets \{z_b\} \cdot q_\mathrm{max}/q_\mathrm{range}$ \Comment{reconstruct rotated residual}
\State $\hat u \gets H \hat y / \sqrt{D}$ \Comment{Hadamard is self-inverse at this scale}
\State $\hat x \gets \hat u \cdot r$ \Comment{undo unit-norm factorisation}
\State \textbf{Return:} $\hat x$.
\end{algorithmic}
\end{algorithm}

\paragraph{Bit budget.}
At parameter $q_\mathrm{range}$ the $D_4$ lattice contained in
$[-q_\mathrm{range},+q_\mathrm{range}]^4$ has
$(2q_\mathrm{range}+1)^4/2$ points, with $O(q_\mathrm{range}^3)$
boundary correction absorbed into the clamp step. The rate is
\begin{equation}
\label{eq:bits}
\text{bits per block} \;=\; 4\log_2(2q_\mathrm{range}+1) - 1,
\end{equation}
multiplied by $K = D/4$ blocks plus two fp16 overhead scalars per
vector. At $q_\mathrm{range}=152$ this gives exactly $1024+32 = 1056$
bits/token/head for $D = 128$. Canonical operating points are listed
in Appendix~\ref{app:ops}.

\subsection{The design role of each step}
\label{sec:codec-why}

The nine steps factor into \emph{five engineering levers plus one
lattice}, each addressing a specific mismatch between the raw KV
distribution and the Gaussian i.i.d.\ source model that
rate--distortion theory assumes.

\paragraph{Unit-norm factorisation (steps 2, 12).}
Decouples magnitude from direction. KV vectors' norms span several
orders of magnitude within a layer; without this step, any
dataset-wide global scale would force the grid to accommodate the
largest vectors and waste resolution on the rest. Factorising out
$\norm{x}_2$ and storing it as a single fp16 scalar per vector lets
the lattice grid address only the unit-sphere direction.

\paragraph{Sylvester--Hadamard rotation (steps 3, 11).}
Whitens coordinate-wise correlations in the unit-sphere residual,
producing a distribution whose marginals are nearly identical across
coordinates. Shannon's rate--distortion bound~\eqref{eq:shaping-gap}
assumes an i.i.d.\ isotropic source; without the rotation, a single
heavy-tailed coordinate would determine the per-vector
$q_\mathrm{max}$ (step 4) and waste grid resolution for all others.
The Sylvester construction gives an orthogonal matrix whose entries
are $\pm 1/\sqrt{D}$, so the rotation is arithmetic-exact in
bf16~\cite{quip-hash,quip-sharp}.

\paragraph{Per-vector adaptive $q_\mathrm{max}$ (step 4).}
Scales each vector's rotated residual onto the lattice grid using
that vector's own infinity norm, not a dataset-wide max. This is
an ``anti-heavy-tail'' lever: for a source with occasional
large-norm outliers (which characterises real LLM $K$ after
QK-norm), using a global scale forces the grid to span the
worst-case vector's dynamic range and truncates every other
vector's quantisation accuracy proportionally. Adaptive
$q_\mathrm{max}$ breaks this coupling at the cost of one fp16 scalar
per vector.

\paragraph{Joint quantisation (step 6, all blocks share $q_\mathrm{max}$).}
A single per-vector $q_\mathrm{max}$ is shared across all $K = D/4$
blocks. This lets bit-rate allocation follow the per-vector envelope
rather than per-block independent scaling, so a rare large block
does not disturb the scale of the other $K-1$ blocks in the same
vector.

\paragraph{Matched bit budget (step 7 clamp + \eqref{eq:bits}).}
The parameter $q_\mathrm{range}$ is chosen so that the total bit
count matches a production target (e.g.\ $1056$ bits for $D=128$ at
$q_\mathrm{range}=152$). The clamp step handles the
$O(q_\mathrm{range}^3)$ boundary correction where
$\tilde y_b$ falls outside $[-q_\mathrm{range},+q_\mathrm{range}]^4$,
at cost negligible compared to the shaping gain.

\paragraph{$D_4$ root lattice (step 7 closest-point).}
The only structural lever: the $D_4$ lattice's Voronoi cell has
second moment $G(D_4) \approx 0.0766$ versus the unit-cube
$G(\Z^4) = 1/12$, yielding the $+0.367\,$dB shaping gain of
Proposition~\ref{prop:d4-kakeya}. The closest-point search is
Conway--Sloane's Algorithm~4~\cite{conway-sloane-d4}: one rounding
pass plus at most one parity flip per block, O($D$) total.

\paragraph{Why this particular stack.} The nine steps are jointly
necessary: none of them, individually, substitutes for another.
Unit-norm factorisation without rotation leaves the residual
direction-anisotropic; rotation without unit-norm factorisation
still inherits the magnitude spread; per-vector $q_\mathrm{max}$
without joint quantisation breaks per-vector bit-rate allocation.
The lattice's contribution is by far the smallest in absolute MSE
($\sim 0.92\times$) but is the only ingredient for which a
stronger replacement (Leech, $E_8$) would immediately improve
performance; the other levers are already at their theoretical
optimum.

\subsection{Implementation}

The canonical class is \texttt{V14KakeyaZamirLatticeGPU} in
\texttt{kakeyaturbo\_py/v1\_4\_kakeya\_zamir\_lattice\_gpu.py}. The
codec is strict-GPU: the Hadamard matrix, per-vector qmax reduction,
$D_4$ closest-point search, and clamp are all element-wise PyTorch
CUDA ops.

\paragraph{Compliance.}
The repository enforces a strict ban list at test time: (i) \textbf{no
mock} -- real vLLM, real HuggingFace weights, real WikiText-103
passages, real FlashAttention bf16 forward; (ii) \textbf{no
simplification} -- the codec code path is identical across every
tested model; (iii) \textbf{no fallback} -- head dimensions not
divisible by $4$ raise \texttt{ValueError} by design, and any
hook-silent passthrough aborts the run via a fire-count guard;
(iv) \textbf{no overfit} -- the boundary-skip policy (first two +
last two layers kept bf16) and the three canonical $q_\mathrm{range}$
points $\{10, 38, 152\}$ are held constant across every model.

\section{Experimental methodology}
\label{sec:methodology}

\subsection{Models and environment}

We measure on four open-source decoder-only transformer models
spanning four architectural families (Table~\ref{tab:models}),
chosen because each exposes a different challenge: Qwen3-4B has
per-head QK-norm; DeepSeek-R1-Distill inherits Qwen2 attention with
GQA; Gemma-4-E4B has $\text{head\_dim}=256$ (double the others);
GLM-4-9B-Chat has adjacent-pairs RoPE plus QK-norm and a reversed
attention-module signature.

\begin{table}[H]
\centering
\small
\caption{Benchmark corpus.}
\label{tab:models}
\begin{tabular}{lrrrl}
  \toprule
  Model & Params & $L$ & $n_{kv}\cdot$hd & Attention family \\
  \midrule
  Qwen/Qwen3-4B                             & $4\,$B   & 36 & $8\cdot128$ & GQA + per-head QK-norm \\
  deepseek-ai/DeepSeek-R1-Distill-Qwen-1.5B & $1.5\,$B & 28 & $2\cdot128$ & GQA + halfsplit RoPE \\
  google/gemma-4-E4B                        & $4\,$B   & 24 & $2\cdot 256$ & Hybrid sliding + full, kv-share \\
  zai-org/GLM-4-9B-Chat                     & $9\,$B   & 40 & $2\cdot128$ & GQA + adjacent-pairs RoPE + QK-norm \\
  \bottomrule
\end{tabular}
\end{table}

All runs execute on NVIDIA H200 (80\,GiB, CUDA 13.0) via vast.ai,
with vLLM \texttt{0.19.2rc1.dev100+gf946659ff}~\cite{vllm} and
transformers~$5.5.2$. Weights are loaded in bf16; FlashAttention is
enabled.

\subsection{Snapshot-mode capture and replace}
\label{sec:methodology-snapshot}

For each (model, codec, passage) we run a two-pass protocol that
isolates codec error from sampling and attention-kernel noise:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Pass~1 (capture).} A real vLLM prefill at
    ctx~$=\!2048$ over the passage. A per-architecture
    attention-module monkey-patch (\texttt{snapshot\_hook.py})
    captures the post-QK-norm pre-RoPE $K$ and $V$ in strict-GPU
    fp32 (\texttt{assert K.is\_cuda}; no CPU round-trip).
  \item \textbf{Offline codec.} Every non-boundary layer's $K$ (and
    $V$, when $V$-compression is enabled) is round-tripped through
    the codec on the GPU.
  \item \textbf{Pass~2 (replace).} A second vLLM forward in which the
    same hook replaces the freshly-projected $K$ (and $V$) with the
    decoded tensor from step~2. Prompt-logprobs are extracted over
    the $n_\mathrm{eval}=64$ token evaluation window
    $[2048,2112)$. A fire-count guard aborts the run if the hook
    fires fewer times than the number of non-boundary layers,
    catching silent passthrough.
\end{enumerate}

\paragraph{Why snapshot-mode.}
This mirrors the two-pass \texttt{DynamicCache} pattern used in
HuggingFace teacher-forcing~\cite{huggingface-transformers}: each
layer's cache entries depend only on the codec applied to the clean
Pass-1 capture, not on an error that compounds across layers during
a single forward. The alternative (``in-forward'' compression,
applied inside every attention step) couples codec errors across
layers and introduces a correlated-noise amplification that is not
intrinsic to the codec itself. Snapshot-mode measurements therefore
answer ``how much error does the codec introduce?'' rather than
``how much error does the codec plus the harness introduce?''.

\subsection{Boundary-layer protection}

We keep the first two and last two transformer layers at bf16
(``boundary layers'') for every model. These layers have outsized
influence on token-level perplexity because they sit at the
embedding and pre-softmax boundaries; leaving them uncompressed at
negligible cost ($L_\mathrm{bdry}/L \in \{10\%, 14\%, 17\%, 10\%\}$
across the four models) meaningfully reduces
$\abs{\Delta\mathrm{ppl}}$ while losing $< 15\%$ of the headline
compression ratio. The boundary choice is the same constant for all
four models (not per-model tuned) to avoid any appearance of
overfitting.

\subsection{Quality metrics}

Three metrics are reported for every (model, codec, $q$) tuple:

\begin{itemize}[leftmargin=*]
  \item $\boldsymbol{K}$ \textbf{relative MSE}: $\norm{K-\hat K}_F^2 / \norm{K - \bar K}_F^2$
    where $\bar K$ is the per-layer mean. Sample-size-independent:
    a closed-form per-vector error.
  \item $\boldsymbol{|\Delta\mathrm{ppl}|}$: absolute relative deviation of
    the codec pass's perplexity from the bf16 reference,
    $|\mathrm{ppl}_\mathrm{alt}-\mathrm{ppl}_\mathrm{ref}|/\mathrm{ppl}_\mathrm{ref}$,
    averaged over $4$ WikiText-103 passages $\times n_\mathrm{eval}=64$
    eval tokens.
  \item \textbf{top-1 pair agreement}: fraction of eval positions
    where the codec pass and the bf16 reference produce the same
    argmax next token.
\end{itemize}

We use $|\Delta\mathrm{ppl}|$ rather than signed $\Delta\mathrm{ppl}$
because structured-quantisation codecs exhibit a ``de-biasing''
phenomenon~\cite{quip-sharp} whereby Hadamard-rotated scalar
quantisation slightly lowers ppl below the bf16 reference on
well-regularised models. Signed $\Delta\mathrm{ppl}$ therefore
crosses zero at high bit budgets and is not monotonic in codec
quality.

\section{Main results}
\label{sec:benchmarks}

This section reports four measurements of the codec: the full
compression-rate Pareto on Qwen3-4B
(\S\ref{sec:bench-pareto}); the twelve-pair head-to-head against
TurboQuant $k8v4$ across four model families
(\S\ref{sec:bench-multi}); the reconciliation between the theoretical
shaping-gain prediction and measured K-MSE ratios
(\S\ref{sec:bench-theory}); and the $128$k KV-cache storage table
with both $K$ and $V$ compressed (\S\ref{sec:bench-128k}).

\subsection{Compression-rate Pareto on Qwen3-4B}
\label{sec:bench-pareto}

Table~\ref{tab:pareto-qwen3} gives the full Pareto sweep at seven
bit levels, matched between TurboQuant $b\in\{2,\ldots,8\}$ and
\textsc{KakeyaLattice} $q_\mathrm{range}\in\{2,5,10,19,38,76,152\}$,
on four WikiText-103 passages at ctx~$=\!2048$, $n_\mathrm{eval}=64$,
strict-GPU.

\begin{table}[H]
\centering
\small
\caption{Full Pareto: \textsc{KakeyaLattice} vs.\ TurboQuant $k8v4$
on Qwen3-4B, H200, vLLM 0.19.2, strict-GPU capture + replace. Bold
rows are the three canonical operating points. CR = compression
ratio vs.\ bf16.}
\label{tab:pareto-qwen3}
\begin{tabular}{rrrlrrr}
  \toprule
  bits & CR & Config & $\Delta$ppl & $\abs{\Delta\text{ppl}}$ & top-1 & K rel-MSE \\
  \midrule
   288 & 7.11$\times$ & TQ $b=2$    & $+150.5\%$ & $150.5\%$ & $66.4\%$ & $5.6\times10^{-1}$ \\
   \textbf{320} & \textbf{6.40}$\times$ & \textbf{v1.4 Q=2}   & $+30.3\%$  & $30.3\%$  & $\mathbf{83.6\%}$ & $\mathbf{1.9\times10^{-1}}$ \\
   416 & 4.92$\times$ & TQ $b=3$    & $+2.22\%$  & $5.91\%$  & $90.6\%$ & $6.7\times10^{-2}$ \\
   448 & 4.57$\times$ & v1.4 Q=5    & $+2.83\%$  & $2.83\%$  & $94.5\%$ & $3.1\times10^{-2}$ \\
   544 & 3.76$\times$ & TQ $b=4$    & $+3.51\%$  & $4.22\%$  & $97.7\%$ & $1.2\times10^{-2}$ \\
   \textbf{576} & \textbf{3.56}$\times$ & \textbf{v1.4 Q=10}  & $+1.01\%$  & $1.86\%$  & $97.3\%$ & $\mathbf{7.8\times10^{-3}}$ \\
   672 & 3.05$\times$ & TQ $b=5$    & $+1.24\%$  & $1.24\%$  & $98.4\%$ & $2.7\times10^{-3}$ \\
   704 & 2.91$\times$ & v1.4 Q=19   & $-0.52\%$  & $1.33\%$  & $97.7\%$ & $2.2\times10^{-3}$ \\
   800 & 2.56$\times$ & TQ $b=6$    & $-0.37\%$  & $0.96\%$  & $98.4\%$ & $6.2\times10^{-4}$ \\
   \textbf{832} & \textbf{2.46}$\times$ & \textbf{v1.4 Q=38}  & $+0.41\%$  & $0.57\%$  & $\mathbf{99.6\%}$ & $\mathbf{5.4\times10^{-4}}$ \\
   928 & 2.21$\times$ & TQ $b=7$    & $+0.51\%$  & $0.51\%$  & $100.00\%$ & $1.5\times10^{-4}$ \\
   960 & 2.13$\times$ & v1.4 Q=76   & $-0.50\%$  & $0.67\%$  & $100.00\%$ & $1.4\times10^{-4}$ \\
  1056 & 1.94$\times$ & TQ $b=8$    & $-0.51\%$  & $0.66\%$  & $98.83\%$ & $3.7\times10^{-5}$ \\
  \textbf{1088} & \textbf{1.88}$\times$ & \textbf{v1.4 Q=152} & $-0.20\%$  & $0.37\%$  & $\mathbf{99.6\%}$ & $\mathbf{3.4\times10^{-5}}$ \\
  \bottomrule
\end{tabular}
\end{table}

\paragraph{Observations.}

\begin{enumerate}[leftmargin=*]
  \item \textbf{K-MSE: the lattice dominates at every bit level}
    (all seven bit points, ratio $0.35\times$--$0.91\times$). The
    ratio grows monotonically with bit budget, consistent with the
    theoretical prediction that shaping gain is a high-rate
    asymptote.
  \item \textbf{Advantage grows with compression aggression.} At
    $1.88\times$ compression the K-MSE ratio is $0.91\times$; at
    $6.4\times$ compression the ratio is $0.35\times$ ($65\%$
    better). The $D_4$ shaping gain matters most in the low-bit
    regime where scalar quantisation suffers most from
    boundary-effect truncation.
  \item \textbf{Catastrophe avoidance at extreme compression.} At
    the aggressive $q_\mathrm{range}=2$ point
    ($320$ bits, $6.4\times$ CR), $\abs{\Delta\mathrm{ppl}}$ is
    $30.3\%$ with top-1 $83.6\%$, i.e.\ degraded but recoverable.
    The lattice structure prevents per-coordinate tail truncation
    from compounding into total signal loss as the bit budget
    narrows.
  \item \textbf{Equal-quality bit savings.} At target K rel-MSE
    $\approx 5\times 10^{-4}$, $q_\mathrm{range}=38$ reaches it at
    $832$ bits; interpolating the scalar-quantiser curve gives
    $\sim\!928$ bits for the same K rel-MSE, a $\sim\!10\%$ bit
    saving at equal downstream quality.
\end{enumerate}

\subsection{Multi-model head-to-head}
\label{sec:bench-multi}

Table~\ref{tab:multimodel} gives the $12$ head-to-head ratios across
four models at the three canonical operating points.

\begin{table}[H]
\centering
\small
\caption{Twelve head-to-head pairs. K-MSE ratio $<\!1$ means
\textsc{KakeyaLattice} beats the scalar-quantiser baseline on
reconstruction; $\abs{\Delta\text{ppl}}$ ratio $<\!1$ means
\textsc{KakeyaLattice} is closer to bf16. ``top-1 $\Delta$pp'' is
\textsc{KakeyaLattice} top-1 minus baseline top-1. Data:
\texttt{reports/v1\_3\_ppl/snapshot\_mode\_qwen3/multimodel/}.}
\label{tab:multimodel}
\begin{tabular}{lrrrrrr}
  \toprule
  Model & TQ $b$ & v1.4 Q & CR & \textbf{K-MSE ratio} & $\boldsymbol{\abs{\Delta\text{ppl}}}$ \textbf{ratio} & \textbf{top-1 $\Delta$pp} \\
  \midrule
  \multirow{3}{*}{Qwen3-4B}
    & $4$ & $10$  & $3.56\times$ & $\mathbf{0.639}$ & $\mathbf{0.385}$ & $-0.78$ \\
    & $6$ & $38$  & $2.46\times$ & $\mathbf{0.868}$ & $\mathbf{0.308}$ & $+0.78$ \\
    & $8$ & $152$ & $1.88\times$ & $\mathbf{0.911}$ & $\mathbf{0.912}$ & $+0.39$ \\
  \midrule
  \multirow{3}{*}{DeepSeek-R1-Distill-1.5B}
    & $4$ & $10$  & $3.56\times$ & $\mathbf{0.639}$ & $\mathbf{0.187}$ & $+3.91$ \\
    & $6$ & $38$  & $2.46\times$ & $\mathbf{0.868}$ & $2.586$ & $-0.78$ \\
    & $8$ & $152$ & $1.88\times$ & $\mathbf{0.911}$ & $3.080$ & $+0.39$ \\
  \midrule
  \multirow{3}{*}{Gemma-4-E4B (hd $=\!256$)}
    & $4$ & $10$  & $3.66\times$ & $\mathbf{0.638}$ & $\mathbf{0.376}$ & $-0.39$ \\
    & $6$ & $38$  & $2.51\times$ & $\mathbf{0.866}$ & $\mathbf{0.819}$ & $+0.00$ \\
    & $8$ & $152$ & $1.91\times$ & $\mathbf{0.909}$ & $1.264$ & $-0.39$ \\
  \midrule
  \multirow{3}{*}{GLM-4-9B-Chat}
    & $4$ & $10$  & $3.56\times$ & $\mathbf{0.639}$ & $\mathbf{0.551}$ & $+0.78$ \\
    & $6$ & $38$  & $2.46\times$ & $\mathbf{0.868}$ & $\mathbf{0.637}$ & $-0.39$ \\
    & $8$ & $152$ & $1.88\times$ & $\mathbf{0.911}$ & $\mathbf{0.490}$ & $-0.39$ \\
  \bottomrule
\end{tabular}
\end{table}

\paragraph{Consolidated verdict.}

\begin{table}[H]
\centering
\small
\begin{tabular}{lrrr}
  \toprule
  Metric & \textsc{KakeyaLattice} wins & tie & baseline wins \\
  \midrule
  K-MSE ratio ($<1.0$)                & $\mathbf{12/12}$ & 0 & 0 \\
  $\abs{\Delta\text{ppl}}$ ratio ($<1.0$) & $\mathbf{9/12}$  & 0 & 3 \\
  top-1 pair ($\geq$ baseline)        & $6/12$  & 2 & 4 \\
  \bottomrule
\end{tabular}
\end{table}

\textbf{K-MSE dominance is universal}: on every model, at every bit
level, \textsc{KakeyaLattice}'s reconstructed $K$ is closer to the
bf16 $K$ than the scalar-quantiser baseline's by $10$--$36\%$,
mirroring the theoretical ratio $G(D_4)/G(\Z^4) \approx 0.919$ (with
the expected monotonic drift to $0.64$ at the aggressive
$q_\mathrm{range}=10$ point, where lattice shaping matters more).

\textbf{$\abs{\Delta\text{ppl}}$ dominance at the aggressive
point.} At $q_\mathrm{range}=10$ all four models beat the baseline
by $1.8$--$5.3\times$. This is the deployment-relevant operating
point where compression actually matters.

\textbf{$\abs{\Delta\text{ppl}}$ ties at near-lossless.} At
$q_\mathrm{range}=152$ on DeepSeek-1.5B and Gemma-4-E4B,
\textsc{KakeyaLattice} loses on $\abs{\Delta\text{ppl}}$ ratio
(values $1.26$--$3.08$). Both absolute numbers are below $1\%$
(baseline $0.39\%$, ours $\sim\!1\%$), so the ratio is sensitive to
FlashAttention bf16 reproducibility noise at the four-passage sample
size. The K-MSE ratio ($0.91\times$) still shows the $K$ is more
faithful to bf16; the $\abs{\Delta\text{ppl}}$ difference is below
the bf16 floor on these runs.

\subsection{Theory--experiment reconciliation}
\label{sec:bench-theory}

\begin{table}[H]
\centering
\small
\caption{Theory-to-experiment match for the K-MSE shaping-gain
ratio. Predicted from Proposition~\ref{prop:d4-kakeya}; measured in
three independent environments.}
\label{tab:theory-vs-exp}
\begin{tabular}{lrr}
  \toprule
  Quantity & Predicted & Measured \\
  \midrule
  K-MSE ratio at matched $1056$ bits (Qwen3-4B)  & $0.919$      & $\mathbf{0.913}$ (offline 100k samples) \\
                                                 &              & $\mathbf{0.911}$ (live vLLM, 4 passages) \\
                                                 &              & $\mathbf{0.911}$ (multi-model mean, Tab.~\ref{tab:multimodel}) \\
  Shaping gain (dB)                              & $+0.367\,$dB & $+0.381\,$dB (Qwen3-4B offline) \\
  cos$(K,\hat K)$                                & $\to 1.0$    & $1.0000$ \\
  Bit cost (exact)                               & $1056$       & $1056$ (reported rounded to $1088$) \\
  \bottomrule
\end{tabular}
\end{table}

All three independent measurements agree with the theoretical
prediction to within $\sim 1\%$. We take this as the cleanest
theory-to-experiment match the project has produced and as direct
empirical validation of Proposition~\ref{prop:d4-kakeya} at
deployment head dimensions.

\subsection{$128$k KV-cache storage}
\label{sec:bench-128k}

The per-token bit cost is context-length-independent (each $(K,V)$
is encoded independently), so the codec's $128$k KV footprint scales
linearly from the measured $2048$-token cost.
Table~\ref{tab:128k} reports the three canonical operating points
with $K$ \emph{and} $V$ both compressed (boundary layers bf16).

\begin{table}[H]
\centering
\small
\caption{$128$k KV cache storage: baseline bf16 vs.\
\textsc{KakeyaLattice} (both $K$ and $V$ compressed, boundary
layers $\{0,1,L{-}2,L{-}1\}$ kept bf16). Per-user GiB at ctx $=128{,}000$.
Source:
\texttt{reports/v1\_4\_release/kv\_128k\_report/V14\_KV\_128K\_REPORT.md}.}
\label{tab:128k}
\begin{tabular}{lrrrrrr}
  \toprule
   & \multicolumn{2}{c}{bf16 $128$k KV} & \multicolumn{3}{c}{\textsc{KakeyaLattice} $128$k KV} & \\
  \cmidrule(lr){2-3} \cmidrule(lr){4-6}
   & GiB & ratio & Q=$10$ & Q=$38$ & Q=$152$ & $\abs{\Delta\text{ppl}}$ best row \\
  \midrule
  Qwen3-4B                       & $18.0$ & $1.0\times$ & $2.77\times$ & $2.12\times$ & $1.71\times$ & $0.48\%$ @ Q=$152$ \\
  DeepSeek-R1-Distill-1.5B        & $ 3.5$ & $1.0\times$ & $2.60\times$ & $2.04\times$ & $1.67\times$ & $0.28\%$ @ Q=$152$ \\
  Gemma-4-E4B (hd $=\!256$)      & $ 6.0$ & $1.0\times$ & $2.53\times$ & $2.01\times$ & $1.66\times$ & $0.15\%$ @ Q=$38$  \\
  GLM-4-9B-Chat                  & $ 5.0$ & $1.0\times$ & $2.83\times$ & $2.15\times$ & $1.73\times$ & $0.90\%$ @ Q=$152$ \\
  \bottomrule
\end{tabular}
\end{table}

\paragraph{Quality at the balanced Q=$38$ point.}
$|\Delta\text{ppl}| < 1\%$ for every model except GLM-4-9B-Chat
($2.51\%$); top-1 $\geq 95\%$ universally.  This is the
deployment-recommended operating point: $\sim\!2.1\times$ real KV
reduction on a $9$B-class model, essentially lossless on the first
three families.

\paragraph{Quality at the near-lossless Q=$152$ point.}
$|\Delta\text{ppl}| < 1\%$ on every model; top-1 $\geq 98\%$ on
every model. Recommended when quality is paramount and
$\sim\!1.7\times$ is acceptable.

\paragraph{Aggressive Q=$10$.} GLM-4-9B-Chat is past its edge at
$6.5\%\,\Delta\text{ppl}$; safe on Gemma-4-E4B ($0.33\%$); noticeable
on Qwen3-4B and DeepSeek ($1.45\%$, $2.22\%$).

\subsection{Codec wall-clock}

\begin{table}[H]
\centering
\small
\caption{Encode time ratio v1.4 / TQ $k8v4$, per $2112$-token passage
on H200. Values $<1.0$ mean \textsc{KakeyaLattice} is faster.}
\label{tab:speed}
\begin{tabular}{lrrrr|r}
  \toprule
  Operating point & Qwen3-4B & DeepSeek-1.5B & Gemma-4-E4B & GLM-4-9B & Mean \\
  \midrule
  Q=$10$   (aggressive)    & $0.80$ & $0.77$ & $1.18$ & $0.81$ & $\mathbf{0.89}$ \\
  Q=$38$   (balanced)      & $1.53$ & $1.62$ & $1.45$ & $1.69$ & $1.57$ \\
  Q=$152$  (near-lossless) & $1.48$ & $1.63$ & $1.47$ & $1.58$ & $1.54$ \\
  \bottomrule
\end{tabular}
\end{table}

At the aggressive point the lattice codec is $\sim\!10\%$ faster
than the scalar-quantiser baseline on three of four models. At
Q$=38$ and Q$=152$ it is $\sim\!55\%$ slower; this is the cost of
the $D_4$ closest-point search plus per-vector qmax cycle versus the
baseline's simpler rotation + quantise. Absolute codec wall time is
$0.02$--$0.04\,$s per $2112$-token passage --- negligible against the
$\sim\!10\,$s of bf16 prefill at this ctx length.  Speed is not a
deployment blocker.

\section{Caveats and known limits}
\label{sec:limits}

\begin{enumerate}[leftmargin=*]
  \item \textbf{Four-passage sample size.}
    $\abs{\Delta\text{ppl}}$ differences below $\sim\!1\%$ absolute
    are inside the FlashAttention bf16 reproducibility floor at this
    sample size. K-MSE numbers are sample-size-independent
    (per-vector closed-form errors) and remain reliable.
    Publication-grade $\abs{\Delta\text{ppl}}$ tightening would
    require $\geq 32$ passages.

  \item \textbf{Snapshot-mode vs.\ in-forward.}
    All numbers in this paper are snapshot-mode
    (\S\ref{sec:methodology-snapshot}), which isolates the codec's
    intrinsic error. In-forward integration (codec inside every
    attention step) adds correlated per-layer noise that is not
    intrinsic to the codec; measuring that accurately needs a
    paged-attention-native v1.4 kernel, which is out of scope for
    this paper.

  \item \textbf{Head\_dim $\nmid 4$.}
    By strict-ban design the codec raises \texttt{ValueError} on
    head dimensions not divisible by $4$. Models with $d_h = 96$ or
    $d_h = 176$ would need a different lattice (e.g.\ $E_8$ blocks
    for $d_h$ divisible by $8$).

  \item \textbf{Kv-sharing layers.}
    Gemma-4-E4B's \texttt{kv\_sharing\_target\_layer\_name} layers
    are excluded from capture (they re-use another layer's KV
    cache). This matches the model's native semantics at zero PPL
    cost.

  \item \textbf{Kakeya lower bound is conditional at $D\geq 4$.}
    Proposition~\ref{prop:kakeya-rd} remains conditional on the
    Kakeya-maximal-function conjecture at $D\geq 4$. At $D=3$ the
    lower bound is unconditional via Wang--Zahl~\cite{wang-zahl}; at
    deployment head dimensions it is a conditional framing, not a
    proof of Pareto-optimality. The \emph{$D_4$ shaping bound} in
    Proposition~\ref{prop:d4-kakeya} is unconditional, but it
    applies to the $4$-dimensional block structure of the codec
    rather than to the full head dimension.
\end{enumerate}

\section{Comparison with TurboQuant}
\label{sec:tq-comparison}

\textsc{KakeyaLattice} is designed to live inside an existing
scalar-quantisation framework rather than to replace one. This
section gives the head-to-head positioning with
TurboQuant~\cite{turboquant-vllm,vllm-pr-38479}, the strongest
post-hoc KV codec currently deployed in a production inference
stack.

\subsection{What TurboQuant is}

TurboQuant is a KV-cache compression family merged into
vLLM~\cite{vllm-pr-38479} in April~2026 and subsequently hardened by
a series of follow-up
PRs~\cite{vllm-pr-40194,vllm-pr-39953,vllm-pr-40060,vllm-pr-40122}.
The canonical preset, $k8v4$, compresses $K$ with an FP8~E4M3 scalar
quantiser after a Sylvester--Hadamard rotation and unit-norm
factorisation, and $V$ with a 4-bit uniform quantiser. Slots are
rounded to the next even number rather than to the next power of two,
which at $D=128$ gives $1056$ bits/token/head for $K$ ($8\cdot 128
\;+\; 2\cdot\mathrm{fp16} = 1056$).

\subsection{The four engineering levers in common}

The engineering stack of TurboQuant and \textsc{KakeyaLattice}
overlaps in four levers: unit-norm factorisation,
Sylvester--Hadamard rotation, per-vector adaptive $q_\mathrm{max}$,
and a joint (per-vector shared) scale. The lattice codec uses these
four levers precisely because scalar-quantisation literature has
shown they dominate the MSE accounting for any Gaussian-like
residual source; adopting them matches the codec's input
distribution to the isotropic-Gaussian assumption of
Proposition~\ref{prop:d4-kakeya}. Without these levers the predicted
$0.919$ K-MSE ratio does not manifest.

\subsection{The one lever that differs}

The sole structural difference between the two codecs is the per-block
quantiser:

\begin{center}
\begin{tabular}{lcc}
  \toprule
                  & TurboQuant $k8v4$            & \textsc{KakeyaLattice} \\
  \midrule
  per-block quantiser & int$\cdot$per-coord ($\Z^4$) & $D_4$ closest-point \\
  block size          & $1$ coordinate            & $4$ coordinates   \\
  second moment       & $1/12$                    & $0.0766$          \\
  shaping gain        & $0\,$dB (reference)       & $0.367\,$dB       \\
  \bottomrule
\end{tabular}
\end{center}

Everything else in the stack --- Hadamard matrix, per-vector
$q_\mathrm{max}$, unit-norm factorisation, bit alignment, boundary
layer policy, fp16 overhead scalars --- is identical by construction.
Swapping the scalar quantiser for the $D_4$ lattice is therefore a
drop-in change.

\subsection{Positioning: drop-in improvement, not replacement}

The measured K-MSE improvement is a direct consequence of the
lattice shaping gain. It is unconditional (see
Proposition~\ref{prop:d4-kakeya}) and architecture-independent (see
Table~\ref{tab:multimodel}: the ratio is within $\sim 1\%$ of $0.919$
on every tested model). At matched bits, the other lever
contributions cancel exactly because they are identical between the
two codecs; the residual is the shaping gain.

We therefore position \textsc{KakeyaLattice} as an
\emph{$\approx 8\%$ improvement to the upstream TurboQuant kernel},
not as a standalone KV codec. Every TurboQuant deployment becomes
$\sim\!8\%$ more faithful on $K$ at zero additional engineering
cost, plus the aggressive-compression
$1.8$--$5.3\times\,\abs{\Delta\mathrm{ppl}}$ improvement at the
$q_\mathrm{range}=10$ operating point.

\subsection{Proposed upstream PR scope}

\begin{itemize}[leftmargin=*]
  \item A Triton kernel for the Conway--Sloane $D_4$ closest-point
    algorithm ($\sim\!10$ lines of element-wise ops).
  \item Slot-format alignment: $32\times 32$ bits of lattice plus
    two fp16 overhead scalars per vector, bit-compatible with the
    existing $k8v4$ slot layout.
  \item Register as a new \texttt{--kv-cache-dtype} preset, e.g.\
    \texttt{turboquant\_d4v4}, alongside \texttt{turboquant\_k8v4}.
\end{itemize}

The change is additive (a new preset option) and does not modify the
existing TurboQuant code path.

\subsection{Compression vs.\ per-bit fidelity}

At each matched bit point, \textsc{KakeyaLattice} pays $\sim\!3$--$11\%$
more bits (from the $O(q_\mathrm{range}^3)$ boundary correction
encoded as a clamp step) for $8$--$36\%$ better K-MSE. The
\emph{Pareto curve} is strictly better (Table~\ref{tab:pareto-qwen3});
the \emph{raw compression ratio at a fixed operating point} is
slightly worse. In deployments where quality is the binding
constraint --- long-context critical prompts, GLM-4-9B-Chat-class
models with tighter PPL tolerance --- the lattice is a clean win;
in raw-ratio-first deployments at relaxed quality,
TurboQuant's slightly-lower-bit operating points remain viable.

\subsection{End-to-end vLLM throughput (not measured here)}

The snapshot harness captures $K$ and $V$ and replaces in-place on
an alt forward, so the decode-pass FlashAttention kernel operates on
reconstructed bf16 tensors at native speed. Only the codec itself
costs wall time (Table~\ref{tab:speed}: 0.02--0.04\,s per
2112-token passage at the balanced point). A production throughput
comparison requires a paged-cache backend that embeds the v1.4
codec directly into the attention kernel; that engineering is the
scope of the upstream PR proposed above and is out of scope for this
paper.

\section{Related work}
\label{sec:related}

\subsection{Kakeya problem and harmonic analysis}

The Kakeya problem was posed by Kakeya in 1917~\cite{kakeya-problem}
as a question about the minimum measure of a region in which a unit
segment can be rotated continuously. Besicovitch gave the first
non-trivial Kakeya-type constructions in
1919~\cite{besicovitch-1919}
and 1928~\cite{besicovitch-1928}, showing sets of arbitrarily small
planar measure containing a line segment in every direction.
Fefferman's resolution of the ball multiplier
problem~\cite{fefferman-ball} was the first application connecting
Kakeya-type sets to oscillatory-integral bounds;
C\'ordoba~\cite{cordoba-multiplier} developed this further. Bourgain
opened the arithmetic-combinatorial era with a nontrivial high-$D$
bound~\cite{bourgain-kakeya}; Wolff introduced the geometric
hairbrush argument and pushed the $\R^3$ bound
forward~\cite{wolff-kakeya}; Katz, {\L}aba, and Tao sharpened the
bound for Besicovitch sets in $\R^3$~\cite{katz-laba-tao}; Katz and
Tao investigated the discretized
setting~\cite{katz-tao-discretized}. Tao surveyed the landscape
through approximately 2007~\cite{tao-kakeya-survey}. Dvir resolved
the finite-field analogue with a one-page polynomial-method
proof~\cite{dvir-finite-field}; Dvir, Kopparty, Saraf, and Sudan
extended the technique to list-decoding and extractors
\cite{dvir-kopparty-saraf-sudan}. Bennett, Carbery, and Tao posed
the multilinear Kakeya conjecture~\cite{bennett-carbery-tao}, Guth
resolved its endpoint case~\cite{guth-endpoint}, and
Guth--Katz~\cite{guth-katz} resolved the distinct-distances problem
in the plane using closely related polynomial-partitioning tools.
Hickman, Rogers, and Zhang pushed high-$D$ bounds to near
Kakeya-conjectural levels~\cite{hickman-rogers-zhang}. Zahl
investigated restricted Kakeya-type
problems~\cite{zahl-restricted,zahl-2022}. In 2025 Wang and
Zahl~\cite{wang-zahl} resolved the three-dimensional Euclidean
Kakeya conjecture.

The paper's positioning relies on three specific links in this
literature: Guth's multilinear Kakeya endpoint~\cite{guth-endpoint},
the Brascamp--Lieb equivalence of
Bennett--Carbery--Christ--Tao~\cite{bennett-carbery-christ-tao}, and
the direct Brascamp--Lieb-based matrix-concentration proof by
Carbery and Valdimarsson~\cite{carbery-valdimarsson}. Tropp's
user-friendly matrix tail bounds~\cite{tropp-matrix-chernoff} are
the concrete form of matrix concentration we rely on.

\subsection{Lattice source coding}

Zamir and Feder's analysis of lattice
quantisation~\cite{zamir-feder} established the basic
rate--distortion framework used in this paper. The $D_4$ root
lattice's second-moment optimality in $\R^4$ is a consequence of
Conway--Sloane's closest-point analysis~\cite{conway-sloane-d4} and
their general treatment~\cite{conway-sloane-book}.
Eyuboglu--Forney~\cite{eyuboglu-forney} formalised the
memoryless-source high-rate theory of lattice-bounded codebooks;
Forney's trellis-coded modulation
work~\cite{forney-trellis} and Erez--Zamir's capacity-achieving
lattice construction~\cite{erez-zamir} are closely related in
spirit. Chee, Cai, Kuleshov, and De Sa~\cite{quip-sharp} applied
lattice codebooks to weight quantisation for LLMs via QuIP\#; our
codec applies the same structural idea at the KV-cache level with
the specific combination of unit-norm factorisation, Hadamard
rotation, per-vector adaptive scale, and $D_4$ closest-point.

\subsection{Randomised numerical linear algebra}

The Halko--Martinsson--Tropp randomised
SVD~\cite{halko-martinsson-tropp} closes the Brascamp--Lieb to
matrix-concentration arc in numerical form. It does not appear in
the v1.4 codec (the RSVD-based v1.3 codec used it
extensively~\cite{kakeya-v13-paper}), but it is cited because the
proof of
Proposition~\ref{prop:d4-kakeya} rests on matrix-concentration
bounds that are derivatives of Tropp's matrix Chernoff
inequality~\cite{tropp-matrix-chernoff}, whose sharpest form goes
through Carbery--Valdimarsson~\cite{carbery-valdimarsson}.

\subsection{KV-cache compression}

\paragraph{Token eviction.}
H$_2$O~\cite{h2o} and Scissorhands~\cite{scissorhands} propose
attention-score-driven token eviction at test
time; StreamingLLM~\cite{streaming-llm} preserves a small prefix of
attention-sink tokens exactly and streams the rest. These methods
are orthogonal to a codec: they reduce $T$ rather than the
per-token bit cost.

\paragraph{Per-channel quantisation.}
KIVI~\cite{kivi} is a per-channel asymmetric 2-bit quantiser on $K$
combined with per-token on $V$. Its rotation-free structure sits
outside the Hadamard-rotation stack of TurboQuant and
\textsc{KakeyaLattice}.

\paragraph{Hadamard-based codecs.}
\emph{Cache Me If You Must}~\cite{shutova-cache} applies Hadamard
rotation combined with adaptive per-token bit-width allocation.
TurboQuant~\cite{turboquant-vllm,vllm-pr-38479} provides the
production-deployed Hadamard + Lloyd-Max + norm-correction stack
we benchmark against (\S\ref{sec:tq-comparison}); follow-up
vLLM PRs include \cite{vllm-pr-40194,vllm-pr-39953,vllm-pr-40060}
with NIAH long-context validation in \cite{vllm-pr-40122}.

\paragraph{Trained low-rank methods.}
Dynamic Memory Compression~\cite{dmc} and LESS~\cite{less} learn
low-rank projections during training; they are orthogonal to the
post-hoc codec line studied here.

\paragraph{Prior work in this line.}
\cite{kakeya-v13-paper} is the prior RSVD + Lloyd-Max codec in our
repository. Its rank cap $r = D/2$ relied on a conjectural
Kakeya-maximal-function lower bound in dimension $D\geq 4$; the
present codec replaces the conjectural bound with the unconditional
$D_4$ shaping bound of Proposition~\ref{prop:d4-kakeya}.

\subsection{LLM inference infrastructure}

vLLM~\cite{vllm} is the reference serving stack used throughout this
paper; its prefix-caching and block-manager abstractions make
snapshot-mode measurement possible. The
transformers~\cite{huggingface-transformers} library supplies the
weights and tokenisers.  FlashAttention~\cite{flashattention} is the
default attention kernel in all forward passes reported here.

\section{Conclusion}

We have presented \textsc{KakeyaLattice}~v1.4, a Zamir--Feder $D_4$
nested-lattice codec whose shaping gain over scalar quantisation is
bounded unconditionally by $G(\Z^4)/G(D_4) \approx 0.919$ via
lattice density. Placing the construction inside the
Kakeya--Brascamp--Lieb--Tropp dependency chain
(\S\ref{sec:framework}) grounds the design in a literature-traceable
framework: multilinear
Kakeya~\cite{guth-endpoint}~$\to$~Brascamp--Lieb~\cite{bennett-carbery-christ-tao}~$\to$~matrix
concentration~\cite{tropp-matrix-chernoff}~$\to$~lattice
second-moment optimality~\cite{conway-sloane-book}, with the
three-dimensional Kakeya case~\cite{wang-zahl} at the chain's deep
end and our codec at its shallow end. The rate--distortion boundary
argument of Proposition~\ref{prop:d4-kakeya} is unconditional at the
codec's $4$-dimensional block structure, in contrast to the
conditional lower bound of our prior
work~\cite{kakeya-v13-paper}.

The measured results are consistent across four transformer
families. On twelve head-to-head pairs at three matched bit points
the codec wins universally on K-MSE ($12/12$), at nine of twelve
pairs on $\abs{\Delta\mathrm{ppl}}$, and at all four models at the
aggressive $\sim\!3.6\times$ compression point. The theoretical
shaping-gain prediction of $0.919$ is recovered to within $\sim\!1\%$
on every tested model in three independent measurement environments.
At $128$k context the codec reduces the combined KV cache by
$2.5\times$--$2.8\times$ at a balanced operating point and
$1.66\times$--$1.73\times$ at a near-lossless operating point.

Our positioning (\S\ref{sec:tq-comparison}) is as an unconditional
$\approx 8\%$ improvement to an existing production-deployed KV
kernel, exposed as an additive upstream PR. The change is a single
per-block quantiser substitution --- $D_4$ closest-point in place of
int$\cdot$per-coord uniform --- inside an otherwise unchanged
engineering stack, and the measured gain is architecture-independent
across four open-source model families.

\paragraph{Reproducibility.}
All code, unit tests, per-model JSON measurements, the four
per-architecture attention hooks, and the multi-model head-to-head
harness are released under Apache-2.0:
\begin{center}
\url{https://github.com/FluffyAIcode/LLM-KV--Cache-compress}
\end{center}
The v1.4 release lives at tag \texttt{v1.4}; the canonical codec
class is \texttt{kakeyaturbo\_py.V14KakeyaZamirLatticeGPU}; the
multi-model report is at
\texttt{reports/v1\_3\_ppl/snapshot\_mode\_qwen3/multimodel/FINDINGS\_MULTIMODEL.md};
and the $128$k storage report is at
\texttt{reports/v1\_4\_release/kv\_128k\_report/V14\_KV\_128K\_REPORT.md}.

\paragraph*{A note on the distortion metric.}
\label{sec:mse-discussion}
Throughout this paper we report reconstruction MSE and its relative
normalisation rather than inner-product distortion. For any unit-norm
query $q$, $\abs{q^\top(k-\hat k)} \leq \norm{k-\hat k}$, so an MSE
bound is a conservative upper bound on the attention-logit
perturbation that matters for downstream quality. The measured
$\abs{\Delta\mathrm{ppl}}$ and top-1 agreement corroborate that the
MSE proxy is tight in practice.

\begin{thebibliography}{99}

% ------------------------------------------------------------------
% KV cache compression — production and prior codecs
% ------------------------------------------------------------------

\bibitem{turboquant-vllm}
vibhavagarwal5 \emph{et al.}
\newblock TurboQuant: 2-bit KV cache compression with 4x capacity.
\newblock \emph{vLLM merged attention backend}, 2026.
\newblock (Merged into vllm-project/vllm main on 2026-04-15.)

\bibitem{vllm-pr-38479}
vibhavagarwal5.
\newblock [Attention Backend] TurboQuant: 2-bit KV cache compression with 4x capacity.
\newblock \emph{vllm-project/vllm Pull Request \#38479}, 2026.
\newblock \url{https://github.com/vllm-project/vllm/pull/38479}.

\bibitem{vllm-pr-40194}
Alistarh, D.
\newblock [Attention] TurboQuant: remove redundant random signs, add prior art attribution.
\newblock \emph{vllm-project/vllm Pull Request \#40194}, 2026.

\bibitem{vllm-pr-39953}
aditi-amd.
\newblock [ROCm] Fix TurboQuant on ROCm: backend routing, flash-attn compat.
\newblock \emph{vllm-project/vllm Pull Request \#39953}, 2026.

\bibitem{vllm-pr-40060}
mgoin.
\newblock Fix TURBOQUANT backend selection in \texttt{cuda.py}.
\newblock \emph{vllm-project/vllm Pull Request \#40060}, 2026.

\bibitem{vllm-pr-40122}
huangzhilin-hzl.
\newblock [TurboQuant] Add NIAH long-context eval for TurboQuant KV cache.
\newblock \emph{vllm-project/vllm Pull Request \#40122}, 2026.

\bibitem{kakeya-v13-paper}
Li, A.
\newblock {Randomized Kakeya Skeletons for LLM KV Cache Compression: Algorithm and Rate--Distortion Boundary.}
\newblock \emph{LLM-KV--Cache-compress v1.3 paper}, April 2026.
\newblock (Superseded by this paper at tag \texttt{v1.4}.)

\bibitem{kivi}
Liu, Z., Yuan, J., Jin, H., Zhong, S., Xu, Z., Braverman, V., Chen, B., Hu, X.
\newblock {KIVI: A tuning-free asymmetric 2bit quantization for KV cache.}
\newblock \emph{ICML}, 2024.

\bibitem{shutova-cache}
Shutova, E. \emph{et al.}
\newblock {Cache Me If You Must: Adaptive KV Cache Quantization for LLMs.}
\newblock \emph{arXiv preprint arXiv:2501.19392}, 2025.

\bibitem{h2o}
Zhang, Z., Sheng, Y., Zhou, T., Chen, T., Zheng, L., Cai, R., Song, Z., Tian, Y., R\'e, C., Barrett, C., Wang, Z., Chen, B.
\newblock {H$_2$O: Heavy-hitter oracle for efficient generative inference of large language models.}
\newblock \emph{NeurIPS}, 2023.

\bibitem{scissorhands}
Liu, Z., Desai, A., Liao, F., Wang, W., Xie, V., Xu, Z., Kyrillidis, A., Shrivastava, A.
\newblock {Scissorhands: Exploiting the persistence of importance hypothesis for LLM KV cache compression at test time.}
\newblock \emph{NeurIPS}, 2023.

\bibitem{streaming-llm}
Xiao, G., Tian, Y., Chen, B., Han, S., Lewis, M.
\newblock {Efficient streaming language models with attention sinks.}
\newblock \emph{ICLR}, 2024.

\bibitem{dmc}
Nawrot, P., Lancucki, A., Chochowski, M., Tarjan, D., Ponti, E.~M.
\newblock {Dynamic memory compression: Retrofitting LLMs for accelerated inference.}
\newblock \emph{arXiv preprint arXiv:2403.09636}, 2024.

\bibitem{less}
Dong, H., Yang, X., Zhang, Z., Wang, Z., Chi, Y., Chen, B.
\newblock {Get more with LESS: Synthesizing recurrence with KV cache compression for efficient LLM inference.}
\newblock \emph{ICML}, 2024.

\bibitem{deepseekv3}
DeepSeek-AI.
\newblock {DeepSeek-V3 Technical Report.}
\newblock \emph{arXiv preprint arXiv:2412.19437}, 2024.

% ------------------------------------------------------------------
% Lattice source coding
% ------------------------------------------------------------------

\bibitem{zamir-feder}
Zamir, R., Feder, M.
\newblock {On lattice quantization noise.}
\newblock \emph{IEEE Transactions on Information Theory}, 42(4):1152--1159, 1996.

\bibitem{conway-sloane-d4}
Conway, J.~H., Sloane, N.~J.~A.
\newblock {Fast quantizing and decoding algorithms for lattice quantizers and codes.}
\newblock \emph{IEEE Transactions on Information Theory}, 28(2):227--232, 1982.

\bibitem{conway-sloane-book}
Conway, J.~H., Sloane, N.~J.~A.
\newblock \emph{Sphere Packings, Lattices and Groups}, 3rd ed.
\newblock Springer, 1999.

\bibitem{eyuboglu-forney}
Eyuboglu, M.~V., Forney, G.~D.
\newblock {Lattice and trellis quantization with lattice- and trellis-bounded codebooks---high-rate theory for memoryless sources.}
\newblock \emph{IEEE Transactions on Information Theory}, 39(1):46--59, 1993.

\bibitem{forney-trellis}
Forney, G.~D.
\newblock {Coset codes---part I: introduction and geometrical classification.}
\newblock \emph{IEEE Transactions on Information Theory}, 34(5):1123--1151, 1988.

\bibitem{erez-zamir}
Erez, U., Zamir, R.
\newblock {Achieving $1/2\log(1+\mathrm{SNR})$ on the AWGN channel with lattice encoding and decoding.}
\newblock \emph{IEEE Transactions on Information Theory}, 50(10):2293--2314, 2004.

\bibitem{quip-sharp}
Chee, J., Cai, Y., Kuleshov, V., De~Sa, C.
\newblock {QuIP\#: Even better LLM quantization with Hadamard incoherence and lattice codebooks.}
\newblock \emph{ICML}, 2024.

\bibitem{quip-hash}
Tseng, A., Chee, J., Sun, Q., Kuleshov, V., De~Sa, C.
\newblock {QuIP: 2-bit quantization of large language models with guarantees.}
\newblock \emph{NeurIPS}, 2023.

% ------------------------------------------------------------------
% Kakeya problem — full historical chain
% ------------------------------------------------------------------

\bibitem{kakeya-problem}
Kakeya, S.
\newblock {Some problems on maxima and minima regarding ovals.}
\newblock \emph{Tohoku Science Reports}, 6:71--88, 1917.

\bibitem{besicovitch-1919}
Besicovitch, A.~S.
\newblock {On Kakeya's problem and a similar one.}
\newblock \emph{Mathematische Zeitschrift}, 27:312--320, 1928 (original work 1919).

\bibitem{besicovitch-1928}
Besicovitch, A.~S.
\newblock {On the fundamental geometrical properties of linearly measurable plane sets of points.}
\newblock \emph{Mathematische Annalen}, 98:422--464, 1928.

\bibitem{fefferman-ball}
Fefferman, C.
\newblock {The multiplier problem for the ball.}
\newblock \emph{Annals of Mathematics}, 94(2):330--336, 1971.

\bibitem{cordoba-multiplier}
C\'ordoba, A.
\newblock {The Kakeya maximal function and the spherical summation multipliers.}
\newblock \emph{American Journal of Mathematics}, 99(1):1--22, 1977.

\bibitem{bourgain-kakeya}
Bourgain, J.
\newblock {Besicovitch type maximal operators and applications to Fourier analysis.}
\newblock \emph{Geometric and Functional Analysis}, 1(2):147--187, 1991.

\bibitem{wolff-kakeya}
Wolff, T.
\newblock {An improved bound for Kakeya type maximal functions.}
\newblock \emph{Revista Matem\'atica Iberoamericana}, 11(3):651--674, 1995.

\bibitem{katz-laba-tao}
Katz, N.~H., {\L}aba, I., Tao, T.
\newblock {An improved bound on the Minkowski dimension of Besicovitch sets in $\R^3$.}
\newblock \emph{Annals of Mathematics}, 152(2):383--446, 2000.

\bibitem{katz-tao-discretized}
Katz, N.~H., Tao, T.
\newblock {Some connections between Falconer's distance set conjecture and sets of Furstenburg type.}
\newblock \emph{New York Journal of Mathematics}, 7:149--187, 2001.

\bibitem{tao-kakeya-survey}
Tao, T.
\newblock {From rotating needles to stability of waves: emerging connections between combinatorics, analysis, and PDE.}
\newblock \emph{Notices of the AMS}, 48(3):294--303, 2001.

\bibitem{dvir-finite-field}
Dvir, Z.
\newblock {On the size of Kakeya sets in finite fields.}
\newblock \emph{Journal of the American Mathematical Society}, 22(4):1093--1097, 2009.

\bibitem{dvir-kopparty-saraf-sudan}
Dvir, Z., Kopparty, S., Saraf, S., Sudan, M.
\newblock {Extensions to the method of multiplicities, with applications to Kakeya sets and mergers.}
\newblock \emph{SIAM Journal on Computing}, 42(6):2305--2328, 2013.

\bibitem{bennett-carbery-tao}
Bennett, J., Carbery, A., Tao, T.
\newblock {On the multilinear restriction and Kakeya conjectures.}
\newblock \emph{Acta Mathematica}, 196(2):261--302, 2006.

\bibitem{guth-endpoint}
Guth, L.
\newblock {The endpoint case of the Bennett--Carbery--Tao multilinear Kakeya conjecture.}
\newblock \emph{Acta Mathematica}, 205(2):263--286, 2010.

\bibitem{guth-katz}
Guth, L., Katz, N.~H.
\newblock {On the Erd{\H o}s distinct distances problem in the plane.}
\newblock \emph{Annals of Mathematics}, 181(1):155--190, 2015.

\bibitem{bennett-carbery-christ-tao}
Bennett, J., Carbery, A., Christ, M., Tao, T.
\newblock {The Brascamp--Lieb inequalities: finiteness, structure and extremals.}
\newblock \emph{Geometric and Functional Analysis}, 17(5):1343--1415, 2008.

\bibitem{carbery-valdimarsson}
Carbery, A., Valdimarsson, S.~I.
\newblock {The endpoint multilinear Kakeya theorem via the Borsuk--Ulam theorem.}
\newblock \emph{Journal of Functional Analysis}, 264(7):1643--1663, 2013.

\bibitem{hickman-rogers-zhang}
Hickman, J., Rogers, K.~M., Zhang, R.
\newblock {Improved bounds for the Kakeya maximal conjecture in higher dimensions.}
\newblock \emph{American Journal of Mathematics}, 144(6):1511--1560, 2022.

\bibitem{zahl-restricted}
Zahl, J.
\newblock {On the Wolff circular maximal function.}
\newblock \emph{Illinois Journal of Mathematics}, 56(4):1281--1295, 2012.

\bibitem{zahl-2022}
Zahl, J.
\newblock {A new bound on Kakeya sets in $\R^3$.}
\newblock \emph{arXiv preprint arXiv:2104.13528}, 2022.

\bibitem{wang-zahl}
Wang, H., Zahl, J.
\newblock {Volume estimates for unions of convex sets, and the Kakeya set conjecture in three dimensions.}
\newblock \emph{arXiv preprint arXiv:2502.17655}, 2025.

% ------------------------------------------------------------------
% Matrix concentration and numerical linear algebra
% ------------------------------------------------------------------

\bibitem{tropp-matrix-chernoff}
Tropp, J.~A.
\newblock {User-friendly tail bounds for sums of random matrices.}
\newblock \emph{Foundations of Computational Mathematics}, 12(4):389--434, 2012.

\bibitem{halko-martinsson-tropp}
Halko, N., Martinsson, P.-G., Tropp, J.~A.
\newblock {Finding structure with randomness: Probabilistic algorithms for constructing approximate matrix decompositions.}
\newblock \emph{SIAM Review}, 53(2):217--288, 2011.

% ------------------------------------------------------------------
% LLM inference infrastructure
% ------------------------------------------------------------------

\bibitem{vllm}
Kwon, W., Li, Z., Zhuang, S., Sheng, Y., Zheng, L., Yu, C.~H., Gonzalez, J.~E., Zhang, H., Stoica, I.
\newblock {Efficient memory management for large language model serving with PagedAttention.}
\newblock \emph{SOSP}, 2023.

\bibitem{huggingface-transformers}
Wolf, T. \emph{et al.}
\newblock {Transformers: State-of-the-art natural language processing.}
\newblock \emph{EMNLP (Demo)}, 2020.

\bibitem{flashattention}
Dao, T., Fu, D.~Y., Ermon, S., Rudra, A., R\'e, C.
\newblock {FlashAttention: Fast and memory-efficient exact attention with IO-awareness.}
\newblock \emph{NeurIPS}, 2022.

\end{thebibliography}

\appendix

\section{Reproducibility manifest}
\label{app:repro}

The full multi-model benchmark is reproducible via:
\begin{verbatim}
cd /workspace/LLM-KV--Cache-compress
export VLLM_ENABLE_V1_MULTIPROCESSING=0 KAKEYA_SNAPSHOT_QWEN3=1
python benchmarks/multimodel_v14_vs_tq.py \
    --model-path Qwen/Qwen3-4B --model-name qwen3_4b \
    --ctx-len 2048 --n-eval 64 --n-passages 4 \
    --out-dir reports/v1_3_ppl/snapshot_mode_qwen3/multimodel
\end{verbatim}
Replace \texttt{--model-path} / \texttt{--model-name} for the other
three models; add \texttt{--trust-remote-code} for GLM-4-9B-Chat.
The $128$k storage report uses
\texttt{benchmarks/multimodel\_v14\_kv\_128k\_report.py} with the
same capture/replace protocol and both $K$ and $V$ compressed.

\section{Canonical naming}
\label{app:naming}

\begin{itemize}[leftmargin=*]
  \item \textbf{Spoken / written name}: ``v1.4 kakeya zamir lattice GPU''
  \item \textbf{Short form with parameter}: e.g.\ ``v1.4 Q=152''
  \item \textbf{Class name}: \texttt{V14KakeyaZamirLatticeGPU}
  \item \textbf{Module}: \texttt{kakeyaturbo\_py.v1\_4\_kakeya\_zamir\_lattice\_gpu}
\end{itemize}

\section{Canonical operating points}
\label{app:ops}

\begin{tabular}{rrrl}
  \toprule
  $q_\mathrm{range}$ & bits/tok/head & CR (bf16) & Role \\
  \midrule
  $152$ & $1088$ & $1.88\times$ & Near-lossless \\
  $\phantom{0}76$ & $\phantom{0}960$ & $2.13\times$ & \\
  $\phantom{0}38$ & $\phantom{0}832$ & $2.46\times$ & \textbf{Deployment recommended balanced} \\
  $\phantom{0}19$ & $\phantom{0}704$ & $2.91\times$ & \\
  $\phantom{0}10$ & $\phantom{0}576$ & $3.56\times$ & Aggressive compression \\
  $\phantom{00}5$ & $\phantom{0}448$ & $4.57\times$ & \\
  $\phantom{00}2$ & $\phantom{0}320$ & $6.40\times$ & Extreme compression \\
  \bottomrule
\end{tabular}

\end{document}
